Joint maximum-likelihood of phylogenies and ancestral states is not consistent

### 2018-08-02 appendix.tex
fix error and add assumption

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
Recall that
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
As we exclude the infinite branch length cases by letting $x_1,y_1,x_2,y_2,w > 0$, we have the strict inequality $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
Thus, regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$.
The remaining variables follow similar arguments.

@@

We now prove the main result.

>>>>

\begin{lemma}
Let $\tau^*=\tau_1$, $t^*=\{x^*, y^*, x^*, y^*, y^*\}$, and $t=\{x_1, y_1, x_2, y_2, w\}$ with $x_1, y_1, x_2, y_2, w > 0$.
Assume $x_1,y_1,x_2 < 1$.
For $0 < x^*, y^* < 1$ and $\boldsymbol\xi_0 = [\emptyset]_{j=1}^\nSiteSplits$, the solution $\hat{t} := \{\hat{x}_1,\hat{y}_1,\hat{x}_2,\hat{y}_2,\hat{w}\}$ given by
\[
\hat{t} = \arg\max_t \ell_{\tau^*, t^*}(\tau_1, t; \boldsymbol\xi_0)
\]
is subject to the constraints $\hat{x}_1,\hat{x}_2 \in (0, x^*]$, $\hat{y}_1 \in (0, y^*]$, and $\hat{y}_2,\hat{w} \in [y^*, 1]$.
\label{lemma:bounds}
\end{lemma}

\begin{proof}
We proceed by analyzing the gradients of each component of the likelihood function.
We first focus on the marginal log-likelihood
\[
\shannonDivergence_{\tau^*,t^*}(\tau_1,t) = \sum_{j=1}^\nSiteSplits \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot \log \Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t).
\]
By Gibbs' inequality, this function attains a unique maximum when
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1, t) = \Pr(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)
\]
for all $j=1,\ldots,\nSiteSplits$ with $t^*=\{x^*,y^*,x^*,y^*,y^*\}$.
We show that this implies $t=t^*$.
Note that exchanging $x_1$ with $x_2$ and $y_1$ with $y_2$ does not affect the marginal log-likelihood, and, as it has a unique maximum, we have $x_1=x_2=x$ and $y_1=y_2=y$ at this maximum.
Referring to the top of Table~\ref{tab:gen-sitepatprob}, we set up simultaneous equations as follows.
The sum of generating probabilities for site splits $\{3\}$ and $\{1,2\}$ yields
\[
2-2(x^*)^2 = 2-2x^2
\]
and for $\{2\}$ and $\{1,2\}$ yields
\[
2-2(y^*)^2 = 2-2y^2.
\]
implying $x=x^*$ and $y=y^*$.
Setting these parameters to their optima, the difference of generating probabilities for site splits $\emptyset$ and $\{1,3\}$ yields
\[
8x^*(y^*)^2 = 8x^*y^*w
\]
implying $w=y^*$.
Thus the unique maximum is $\hat{t}=\{x^*,y^*,x^*,y^*,y^*\}$, and consequently
\[
\nabla_t \shannonDivergence_{\tau^*,t^*}(\tau_1,t) \mid_{t=t^*} = 0.
\]
We now show that the gradient of the marginal log-likelihood is a decreasing function on $(0,1]$ in each variable.
The marginal log-likelihood as a function of $x_1$ is
\begin{align*}
\shannonDivergence(x_1) = \sum_j p_{\siteSplit_j} \log(c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1) - \log 8
\end{align*}
where $c_{x_1,\siteSplit_j}$ and $a_{x_1,\siteSplit_j}$ are functions of $y_1,x_2,y_2,$ and $w$.
For example, reading from Table~\ref{tab:gen-sitepatprob},
\[
c_{x_1,\emptyset} = 1+y_1y_2+wx_2y_1+wx_2y_2
\]
and
\[
a_{x_1,\emptyset} = x_2+wy_1+wy_2+x_2y_1y_2
\]
with similar definitions for the remaining site splits.
By the assumption of $0 < x^*, y^* < 1$, we have $p_{\siteSplit_j} > 0$ for $j=1,\ldots,\nSiteSplits$, therefore we assume $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 \neq 0$ for all $j$ otherwise we obtain a zero value for our likelihood, i.e., $\shannonDivergence\rightarrow -\infty$.
Assuming $x_1,y_1,x_2 < 1$ precludes all cases where $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 = 0$ for some $j$.
Moreover, we have
\[
\Pr(\siteSplitRV=\siteSplit_j \mid \tau_1,\{x_1,y_1,x_2,y_2,w\}) = \frac{1}{8} (c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1),
\]
which, being a probability, is greater than or equal to zero.
Consider the derivative with respect to $x_1$
\begin{align*}
\frac{d}{dx_1} \shannonDivergence(x_1) = \sum_j \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}
\end{align*}
and focus on the $j$th term in the summation
\[
f_j(x_1) := \frac{p_{\siteSplit_j}a_{x_1,\siteSplit_j}}{c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1}.
\]
%das: gah, had to roll back this line and add some stuff above since there are cases where some probabilities are 0. Maybe I'm being too careful? I don't think it's an issue, just an edge case that could induce wonkiness, so I included an assumption.
Regardless of $a_{x_1,\siteSplit_j} \lessgtr 0$, $f_j$ is a decreasing function of $x_1$ as $c_{x_1,\siteSplit_j} + a_{x_1,\siteSplit_j}x_1 > 0$.
The remaining variables follow similar arguments.

@@

Since $x^*, y^* < 1$, the assumption that $x_1,y_1,x_2 < 1$ is reasonable.
We now prove the main result.
